Future Cosmic Microwave Background delensing with galaxies surveys.

### 2017-06-21 Notes/galaxy_delensing.tex

The measurement of the cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes is indeed a direct signature of the presence of gravitational waves in the early universe. However improving the level of noise is not enough. 
This is even truer if the aim is to not only detect the amplitude of gravitational waves but also the shape of their spectrum to test for example inflation consistency relations.
Removing the lensing component from the measurement of CMB B-modes will be important to constrain the amplitude of the primordial gravitational wave contribution to the signal.
Here we discuss the role of current and future large scale structure surveys in improving the reconstruction of the lensing potential that lenses the CMB photons and how this reflects into an improved delensing efficiency.
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$).
We find that current generation experiments... we also point out that a tomographic approach..
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
In the standard cosmological paradigm, the universe underwent a period of near-exponential expansion in its early phase;
this period is called ``cosmic inflation.''
Inflation generically predicts a stochastic background of gravitational waves \cite[see e.g.][for a review]{kamionkowski15}.
These primordial gravitational waves in the early universe imprint a unique signature on the polarized anisotropies of the CMB. For this reason, CMB polarization is a promising dataset to understand the physics of the early universe.
Furthermore, CMB polarization is usually studied in Fourier space by decomposing the fields into even-parity (divergence) and odd-parity (curl) components, referred to as ``E'' and ``B'' modes.
The B-mode polarization, in particular, is a great source of information because, in the standard scenario, the PGW background is the only source of B-modes at the epoch of recombination.

To connect data and theoretical models, the amplitude of this primordial B-mode component is parameterized by the ratio of the amplitudes of the primordial tensor and scalar perturbations, $r$. The shape of this component, i.e. the amplitude as a function of the angular scales is encoded in the tensor tilt $n_{t}$.
Most of the inflationary models directly relate the value of r to the energy scale of inflation. Measuring or constraining the value of $r$ and $n_t$ will put a tight constraint on the allowed models of inflation.

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational waves background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, a possibly natural value $r \gtrsim 10^{-3}$, should be reachable given the level of noise expected in future CMB experiments \citep{kamionkowski15}.

However reducing the level of noise will not be enough to reach this goal.
Indeed the observed B modes are not solely due to primordial gravitational waves but also by secondary effects taking place in the late-time low-redshift universe. In this work, we will focus on the effects of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,polarbear2014c,vanengelen14b,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

This component acts as a source of confusion for searches of the PGW background.
Indeed, the contamination from lensing B modes is already at the level of the instrumental noise of current experiments \citep{bk14}. Thus together with the experimental effort to reduce the amount of noise is important to study the effect of the spurious lensing component.  
The optimal way to reduce the lensing contributions is to reconstruct the expected realization of lensing B modes on the observed part of the sky and then use it to clean the data in a process called ``delensing''. 
We can delens the observed B-modes combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.

Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, recently it has been performed on CMB temperature data using as LSS tracers the cosmic infrared background in \cite{larsen:2016} and on CMB temperature and polarization data using internal CMB lensing reconstruction \cite{ carron17}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in where $~28\%$ of the lensing power has been removed.

In this paper, we study the possible role of future galaxies surveys as tracers of the lensing potential. These will translate into a better reconstruction of the B-modes in the measured patch, and as a consequence, it will improve the constraints on inflation trough delensing. We model several actual and future surveys, and after computing the residual B-modes, we forecast the statistical uncertainness on $r$ and $n_T$ for current CMB experiments as well as the next generation (S3) and the planned fourth generation (S4).

@@

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light and $P(k,z)$ is the matter power spectrum evaluated at wavenumber $k = \ell/\chi(z)$ and redshift $z$. 
Furthermore $W^{i}(z)$ is the kernel function of the field $i$ introduced in \refeq{wkernel}.
We will now introduce the kernels for each of the tracers used in this work.

@@

The galaxies clustering kernel is:
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies observed by the survey as a function of redshift while $b(z)$ is the galaxy bias that connects the amplitude of galaxies overdensity to the one of the dark matter underneath.
When computing the auto-spectrum of the galaxies density, a shot noise term needs to be taken into account. 
To do so, we add a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.
 Different galaxy surveys in this work are fully characterized by their b(z), $\frac{dN}{dz}$ and the observed galaxies density.
We test the delensing efficiency taking into account both current survey like WISE or DES as well as future galaxy probes like DESI and LSST together with 21 cm measurement like SKA.

@@

However, this promising sign of primordial gravitational waves is partially obscured by gravitational interaction with large scale structure that generates CMB B-modes by distorting primordial E-modes.
At first order, given the convergence field $\kappa= -\frac{1}{2}\nabla^2\phi$ introduced in \refsec{kappaCMB} the B-modes resulting from the lensing of primordial E mode are:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}).
\ee
As usual we define the power spectrum as:
\be
\langle B^{\mathrm{lens}}(\bl) ~ B^{\mathrm{lens}^*}(\tilde{\bl}) \rangle \equiv (2\pi)^2 \delta^D(\bl - \tilde{\bl}) C^{BB,\mathrm{lens}}_l .
\ee

From this we get that the power spectrum of the lensing component of the B-modes:
\be
C_l^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The full B-mode power spectrum measured on the sky is given by a possible primordial component $C_l^{BB,r}$ together with the lensing $C_l^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_l^{BB}$:
\be
C_l^{BB,\mathrm{full}} = C_l^{BB,r} + C_l^{BB,\mathrm{lens}} + N_l^{BB}.
\ee
Given the current constraints, the lensing component is a significant source of B-modes that, at large scales, correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the allowed inflationary component at scales smaller than several degrees, but it is also comparable with current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we can build a template \refeq{blens} of the actual lensing B-modes given a measurement of the E-mode field and the lensing potential $\phi$ in the observed patch.
While E is measured directly, we ca estimate $\phi$ using "tracers" of the dark matter distribution that creates the potential.

We will now show how the delensing efficiency is related to the fidelity of the lensing tracers and instrumental noise in the CMB E-modes.
If we have a large scale structure field I($\nver$) that traces the lensing potential responsible for the lensing of the CMB we can build a template of the lensing B-modes on the sky by a weighted convolution:
\be
\hat{B}^{\mathrm{lens}}(\bl) = \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') f(\bl,\bl') E^N(\ublu) I(\bl - \bl'),
\ee
where $f(\bl,\bl')$ can be determined by minimizing the difference with the true $B^{\mathrm{lens}}(\bl)$ defined in \refeq{blens}. We introduce the instrumental noise in the CMB E-modes ($E^N$) that will also limits the ability to fully reconstruct the lensing B-modes.

@@

The weights $f(\bl,\bl')$ so that the residual lensing B mode power is minimized are:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum and are described for each LSS field in \refsec{model}.
Notice that the first term consists of an inverse variance filter applied to the measured E-mode and the second minimize the difference between the reconstructed $\phi$ and the CMB lensing potential.

With this choice of $f(\bl,\bl')$ we finally have that the residual power is:
\beqn
\label{eqn:Bres}
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be
\rho_l= \frac{\cl^{\kappa I}}{\sqrt{\cl^{\kappa \kappa} \cl^{I I}}}.
\ee
The bigger the $\rho_l$ is for an LSS field, the more it is correlated with the lensing potential acting on the CMB photons. A higher correlation allows for a better reconstruction of the $\phi^{CMB}$ and, as a consequence, of $B^{lens}$.

@@

In this section we extend the formalism to the case where multiple tracers are used to reconstruct the lensing potential. Furthermore we show that splitting galaxy surveys into multiple tomographic bins can significantly improve their efficiency.

@@

Note that the gain in adding a new tracer is not only proportional to its correlation with the CMB lensing, but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in \refsec{model}.
Galaxies clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST, DES and DESI curves.
However given the significant low noise of this measurement they can still help the delensing of the CMB.
On the other end, CIB and 21 cm surveys are probing to high redshift structures and independently from the model assumed they show a relatively good overlap with the CMB lensing kernel.

\subsection{Improving efficiency with tomographic binning}\label{sec:tomo-bin}
When we weight a tracer with $\frac{C^{\kappa I}}{C^{II}}$ in \refeq{fweight} to maximize its ability to reconstruct the lensing potential we are only using redshift averaged information about the survey. However, as can be seen in \reffig{kernel}, the overlap varies as a function of redshifts. For this reason, the optimal approach is to weight each redshift bin with different weights according to both their cross-correlation with $\kappa$ and their auto-spectrum.
We can see this with a simple example. Let's split a single survey I into two non-overlapping redshift bins $I_{1}$ and ${I_{2}}$ with $I=I_{1}+I_{2}$.
For the full survey the effective cross correlation is equal to 
\be
\rho_{\rm{full}}^{2}= \frac{(\cl^{\kappa I})^2}{\cl^{\kappa \kappa} \cl^{I I}}=\frac{(\cl^{\kappa I_{1}}+\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} (\cl^{I_{1} I_{1}}+\cl^{I_{2} I_{2}})}
\ee
while for the split survey it will be 
\be
\rho_{\rm{split}}^{2}= \frac{(\cl^{\kappa I_{1}})^2}{\cl^{\kappa \kappa} \cl^{I_{1} I_{1}}}+\frac{(\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} \cl^{I_{2} I_{2}}}.
\ee
Now it can be show that $\rho_{\rm{split}}\ge \rho_{\rm{full}}$ being:
\be
\rho_{\rm{split}}^{2} - \rho_{\rm{full}}^{2}= 
\ee
The two are equal only when $\frac{C^{\kappa I_{i}}}{C^{I_{i}I_{i}}}$ is the some for all the redshift bins in which case a single optimal weight is sufficient for the entire survey.
Of course, this is an idealized situation: increasing the number of bins will reduce the number of galaxies in the bin thus increasing the shot noise. 
An optimal balance needs to be found given the survey specifics, but the general idea is that the optimal approach consists in binning the galaxies in the survey at different redshift. 
Of course, this is possible for galaxies surveys, but it is harder for a tracer with no redshift information like the CIB.

@@

@@

\subsection{Fisher method}
In this framework, the error on a cosmological parameter constrained by a B-mode spectrum measurement is given as usual by:
\begin{equation}
\label{eqn:fisher}
\sigma(p)= \left[\sum_l^{\ell_{\rm{max}}} \left(\frac{\partial {C}_{\ell}^{BB,r,n_{t}}}{\partial p}\right)^2/\mathrm{Cov}^2( C_{\ell}^{BB} ) \right]^{-\frac{1}{2}}
\end{equation}
where we consider the parameters $p =\{r,n_{t}\}$. 
We will assume a gaussian covariance:
\beq
\sigma( C_{l}^{BB} )= \sqrt{ \frac{2}{(2 {l}+1) \fsky }} \left( C_l^{BB,r} + C_{{l}}^{BB,\mathrm{lens}}+N_{{l}}^{BB} \right).
\eeq
Here the B-modes noise spectrum is given by:
\beq
N_l^{BB} = \left({\Delta_P}/{T_{\mathrm{CMB}}}\right)^2  e^{ {l^2 \theta_\mathrm{FWHM}^2}/({8 \ln 2})}
\eeq
where $\theta_\mathrm{FWHM}$ is the full half width of the telescope beam and $\Delta_P$ is the instrumental noise of the experiment.
Even if the covariance structure has non-Gaussian contributions \cite{motloch:2017,benoit-levy:2012} this approximation is good enough to show the improvement due to delensing.

@@

\bibliographystyle{plain}
\bibliography{DES_rec_phi_paper,/Users/alessandromanzotti/Work/Astrophysics/latex_bib/cosmobib,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/2016/delens100d/delens100d,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}

>>>>

The measurement of the cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However improving the level of noise in future experiments will not be enough. 
%This is even truer if the aim is to not only detect the amplitude of gravitational waves but also the shape of their spectrum to test for example inflation consistency relations.
Secondary effects in the low redshift universe will also produce non-primordial B-modes. In particular the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing component from the measurement of CMB B-modes will be important to constrain the amplitude of the primordial gravitational wave contribution to the signal.
Here we discuss the role of current and future large scale structure surveys in improving the reconstruction of the lensing potential that lenses the CMB photons and how this reflects into an improved delensing efficiency.
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$).
We find that current generation experiments... we also point out that a tomographic approach..
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
In the standard cosmological paradigm, in its early phase, the universe underwent a period of near-exponential expansion called "cosmic inflation." 
So far all the cosmological observations agree with this picture, making it a compelling and elegant description of the Universe initial conditions. Despite the experimental effort, other possible explanations are still valid, and a conclusive evidence of inflation is still to be found. 
Inflation generically predicts a stochastic background of gravitational waves \cite[see e.g.][for a review]{kamionkowski15}. This prediction set inflation apart from other theories and a detection of primordial gravitational waves could be the compelling evidence cosmologist are looking for. 
These primordial gravitational waves in the early universe would imprint a unique signature on the polarized anisotropies of the CMB. For this reason, CMB polarization is a promising dataset to understand the physics of the early universe and ultimately test inflation.
In particular, if we can decompose the CMB polarization fields in Fourier space into even-parity (divergence) and odd-parity (curl) components, referred to as ``E'' and ``B'' modes.
 In the standard scenario, the B-mode polarization is a clean probe of primordial gravitational waves, because these are the only source of B-modes at the epoch of recombination.

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational waves background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, a natural value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski15}.
However, reducing the level of noise will not be enough to attain this goal.
Indeed the observed B modes are not solely sourced by early universe physics, but they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effects of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,polarbear2014c,vanengelen14b,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

This component acts as a source of confusion for searches of the primordial gravitational waves background.
Indeed, the contamination from lensing B modes is already at the level of the instrumental noise of current experiments \citep{bk14}. Thus together with the experimental effort to reduce the amount of noise is important to study the effect of the spurious lensing component.

The optimal way to reduce the lensing contributions is to reconstruct the expected realization of lensing B modes on the observed part of the sky and then use it to clean the data in a process called ``delensing''. 
We can delens the observed B-modes combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.
Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, recently it has been performed on CMB temperature data using as LSS tracer the cosmic infrared background in \cite{larsen:2016} and on CMB temperature and polarization data using internal CMB lensing reconstruction \cite{ carron17}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in where $~28\%$ of the lensing power has been removed.

For future experiments, we need to increase the delensing efficiency by almost a factor of 3 to fully exploit the instrumental capabilities \cite{abazajian:2016}.
In this paper, we propose and study a possible way: using future galaxies surveys as tracers of the lensing potential in addition to other probes like 21 cm, cosmic infrared background and internal CMB reconstruction. 
Furthermore, we point out how using redshift information trough a clever tomographic binning can improve the delensing efficiency of galaxies survey.
These will translate into a better reconstruction of the B-modes in the measured patch, and as a consequence, it will improve the constraints on inflation trough delensing. We model several actual and future surveys, and after computing the residual B-modes, we forecast the statistical uncertainties the amplitude and the shape of the inflationary tensor perturbations for current CMB experiments as well as the next generation (S3) and the planned fourth generation (S4).

@@

In this equation, $H(z)$ is the Hubble factor at redshift $z$, $c$ is the speed of light and $P(k,z)$ is the matter power spectrum evaluated at wavenumber $k = \ell/\chi(z)$ and redshift $z$. 
Furthermore $W^{i}(z)$ is the kernel function of the field $i$ introduced in \refeq{wkernel}.
We will now describe the kernels for each of the tracers used in this work.

@@

The galaxies clustering kernel is:
\begin{equation}
\label{eqn:wg}
\begin{split}
W^{g}(z) &= \frac{b(z)\frac{dN}{dz}}{\Bigl(\int dz'\,\frac{dN}{dz'}\Bigr)}.
\end{split}
\end{equation}
Here $\frac{dN}{dz}$ is the number of galaxies observed by the survey as a function of redshift while $b(z)$ is the galaxy bias that connects the amplitude of galaxy overdensities to the ones of the dark matter underneath.
When computing the auto-spectrum of the galaxies density, a shot noise term needs to be taken into account. 
To do so, we add a constant term to the power spectrum equal to the inverse of the number of galaxies per steradians.
 Different galaxy surveys in this work are then fully characterized by their b(z), $\frac{dN}{dz}$ and the observed galaxies density.
We test the delensing efficiency taking into account both current survey like WISE or DES as well as future galaxy probes like DESI and LSST together with 21 cm measurement like SKA.

@@

However, this promising sign of primordial gravitational waves is partially obscured by gravitational interactions with large scale structures that generate CMB B-modes by distorting primordial E-modes.
At first order, given the convergence field $\kappa= -\frac{1}{2}\nabla^2\phi$ introduced in \refsec{kappaCMB} the B-modes resulting from the lensing of primordial E-modes are:
\be
B^{\mathrm{lens}}(\bl) =  \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') E(\ublu) \kappa(\bl - \bl')
\label{eqn:blens}
\ee
where different modes contributes with a different weight:
\be
W(\bl,\bl') = \frac{2 \bl' \cdot (\bl-\bl')}{|\bl-\bl'|^2} \sin(2\varphi_{\bl,\bl'}).
\ee
%As usual we define the power spectrum as:
%\be
%\langle B^{\mathrm{lens}}(\bl) ~ B^{\mathrm{lens}^*}(\tilde{\bl}) \rangle \equiv (2\pi)^2 \delta^D(\bl - \tilde{\bl}) C^{BB,\mathrm{lens}}_l .
%\ee

From this we get the power spectrum of the lensing component of the B-modes:
\be
C_l^{BB,\mathrm{lens}}  = \int \frac{d^2 \bl'}{(2 \pi)^2} W^2 (\bl,\bl')C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}.
\ee

The B-mode power spectrum measured on the sky is composed of a possible primordial component $C_l^{BB,r}$ together with the lensing $C_l^{BB,\mathrm{lens}}$ contribution and the instrumental noise $N_l^{BB}$ (defined in \refeq{noise}):
\be
C_l^{BB,\mathrm{full}} = C_l^{BB,r} + C_l^{BB,\mathrm{lens}} + N_l^{BB}.
\ee
The lensing component is a significant source of B-modes that, at large scales, correspond to a white noise source of roughly $5 \mu K$-arcmin. This means that it is not only bigger than the allowed inflationary component at scales smaller than several degrees, but it is also comparable with current levels of instrumental noise. 
For this reason, it is critical to characterize and eventually remove it from the data.
To do so, we build a template \refeq{blens} of the lensing B-modes in the observed patch given a measurement of the E-mode field and the lensing potential $\phi$.
While E is measured directly, we can estimate $\phi$ using "tracers" of the dark matter distribution that sources the potential.

We will now show how the delensing efficiency is related to the fidelity of the lensing tracers and the instrumental noise in the CMB E-modes.
If we have a large scale structure field I($\nver$) that traces the lensing potential we can build a template of the lensing B-modes on the sky by a weighted convolution:
\be
\hat{B}^{\mathrm{lens}}(\bl) = \int \frac{d^2 \bl'}{(2 \pi)^2} W(\bl,\bl') f(\bl,\bl') E^N(\ublu) I(\bl - \bl'),
\ee
where $f(\bl,\bl')$ can be determined by minimizing the difference with the true $B^{\mathrm{lens}}(\bl)$ defined in \refeq{blens}. We introduce the instrumental noise in the CMB E-modes ($E^N$) that will also limits the ability to fully reconstruct the lensing B-modes.

@@

The weights $f(\bl,\bl')$ so that the residual lensing B mode power is minimized are:
\be
\label{eqn:fweight}
f(\bl,\bl') = \left(\frac{C^{EE}_{{l'}}}{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right)  \frac{C^{\kappa I}_{|\bl-\bl'|}}{C^{II}_{|\bl-\bl'|} }.
\ee
Here $C^{\kappa I}$ and $C^{II}$ are the cross correlation spectrum of the tracer $I$ with the lensing convergence $\kappa$ and its autospectrum and are described for each LSS field in \refsec{model}. The power spectrum of the E-modes noise $N^{EE}$ is the same as the B-modes one in \refeq{noise}.

Notice that the first term consists of an inverse variance filter applied to the measured E-mode and the second minimize the difference between the reconstructed $\phi$ and the CMB lensing potential.

With this choice of $f(\bl,\bl')$ we finally have that the residual power is:
\beqn
\label{eqn:Bres}
C_{{l}}^{BB,\mathrm{res}} &=&   \int \frac{d^2 \bl'}{(2 \pi)^2}  W^2 (\bl,\bl')
 C^{EE}_{l'} C^{\kappa \kappa}_{|\bl-\bl'|}  \\
&\times&  \left[1 - \left(\frac{C^{EE}_{{l'}} }{C^{EE}_{{l'}}+N^{EE}_{{l'}}}\right) \rho^2_{|\bl-\bl'|} \right] \nonumber
\eeqn
with
\be
\rho_l= \frac{\cl^{\kappa I}}{\sqrt{\cl^{\kappa \kappa} \cl^{I I}}}.
\ee
The bigger the $\rho_l$ is for an LSS field, the more it is correlated with the lensing potential acting on the CMB photons. An higher correlation allows for a better reconstruction of $\phi$ and, as a consequence, of $B^{\rm{lens}}$.

@@

In this section we extend the formalism to the case where multiple tracers are used to reconstruct the lensing potential. 
%Furthermore we show that splitting galaxy surveys into multiple tomographic bins can significantly improve their efficiency.

@@

Note that the gain in adding a new tracer is not only proportional to its correlation with the CMB lensing, but it also depends on how much it is correlated with the already used set of tracers.
\reffig{kernel} show the different kernels as a function of redshift computed using the models and parameters described in \refsec{model}.
CIB and 21 cm surveys probes the high redshift structures and, independently from the model assumed, they show a relatively good overlap with the CMB lensing kernel. 
On the other end, galaxies clustering surveys can only reconstruct the low-z portion of the lensing kernel as can be seen from the LSST, DES and DESI curves.
However, given the low noise of this measurement and their small overlap with other probes they still improve the delensing efficiency even if they overlap with the CMB lensing potential is not optimal.

\subsection{Improving efficiency with tomographic binning}\label{sec:tomo-bin}
Furthermore the delensing efficiency of galaxy surveys can be improved by taking into account redshift information.
When we weight a tracer with $\frac{C^{\kappa I}}{C^{II}}$ in \refeq{fweight} to maximize its ability to reconstruct the lensing potential we are only using redshift averaged information about the survey. However, as can be seen in \reffig{kernel}, the overlap varies as a function of redshifts. For this reason, the optimal approach is to weight galaxies at different redshifts with different weights according to both their cross-correlation with $\kappa$ and their auto-spectrum.
We can see this with a simple example. Let's split a single survey I into two non-overlapping redshift bins $I_{1}$ and ${I_{2}}$ with $I=I_{1}+I_{2}$.
For the full survey the effective cross correlation is equal to 
\be
\rho_{\rm{full}}^{2}= \frac{(\cl^{\kappa I})^2}{\cl^{\kappa \kappa} \cl^{I I}}=\frac{(\cl^{\kappa I_{1}}+\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} (\cl^{I_{1} I_{1}}+\cl^{I_{2} I_{2}})}
\ee
while for the split survey it will be 
\be
\rho_{\rm{split}}^{2}= \frac{(\cl^{\kappa I_{1}})^2}{\cl^{\kappa \kappa} \cl^{I_{1} I_{1}}}+\frac{(\cl^{\kappa I_{2}})^2}{\cl^{\kappa \kappa} \cl^{I_{2} I_{2}}}.
\ee
Now it can be show that $\rho_{\rm{split}}\ge \rho_{\rm{full}}$ being:
\be
\rho_{\rm{split}}^{2} - \rho_{\rm{full}}^{2}\propto  (\cl^{I_{1} I_{1}}\cl^{\kappa I_{2}}-\cl^{I_{2} I_{2}}\cl^{\kappa I_{1}})^2
\ee
Then $\rho^2$ is always bigger in the tomographic case and the two are equal only when $\frac{C^{\kappa I_{i}}}{C^{I_{i}I_{i}}}$ is the some for all the redshift bins in which case a single optimal weight is sufficient for the entire survey.
Of course, this is an idealized situation: increasing the number of bins will reduce the number of galaxies in the bin thus increasing the shot noise. 
An optimal balance needs to be found given the survey specifics, but the general idea is that the optimal approach consists in binning the galaxies in the survey at different redshift. 
Of course, this is possible for galaxies surveys, but it is harder for a tracer with poor redshift information like the CIB.

@@

\textbf{ .. add binning used ...}

@@

\subsection{Fisher method}
In this framework, the error on a cosmological parameter constrained by a B-mode spectrum measurement is given as usual by:
\begin{equation}
\label{eqn:fisher}
\sigma(p)= \left[\sum_l^{\ell_{\rm{max}}} \left(\frac{\partial {C}_{\ell}^{BB,r,n_{t}}}{\partial p}\right)^2/\sigma( C_{l}^{BB} )^2 \right]^{-\frac{1}{2}}
\end{equation}
where we consider the parameters $p =\{r,n_{t}\}$. 
We will assume a gaussian covariance:
\beq
\sigma( C_{l}^{BB} )= \sqrt{ \frac{2}{(2 {l}+1) \fsky }} \left( C_l^{BB,r} + C_{{l}}^{BB,\mathrm{lens}}+N_{{l}}^{BB} \right).
\eeq
Here the B-modes noise spectrum is given by:
\beq \label{eqn:noise}
N_l^{BB} = \left({\Delta_P}/{T_{\mathrm{CMB}}}\right)^2  e^{ {l^2 \theta_\mathrm{FWHM}^2}/({8 \ln 2})}
\eeq
where $\theta_\mathrm{FWHM}$ is the full half width of the telescope beam and $\Delta_P$ is the instrumental noise of the experiment.
Even if the covariance structure has non-Gaussian contributions \cite{motloch:2017,benoit-levy:2012} this approximation is good enough to show the improvement due to delensing.

@@

\bibliographystyle{plain}
\bibliography{/Users/alessandromanzotti/Work/Astrophysics/latex_bib/cosmobib,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/2016/delens100d/delens100d,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}
